---
title: 'Birds 00: Spatial Analysis'
author: "Arne Wenzel"
date: "June 2019"
output:
  html_document:
    df_print: paged
    toc: yes
---

### Prerequisites

```{r  message=FALSE, warning=FALSE}
library(raster)
library(rgdal)
library(landscapemetrics)
library(dplyr)
library(vegan)
library(tidyverse)
library(berryFunctions)
library(cowplot)
library(ggplot2)
```

### Import raster

The raster I import is a map of the entire Bangalore study region. It does contain both the north and the south transect. However, at the moment 2 of my sites fall outside of the classification.

```{r results='hide'}
bgl <- raster(readGDAL("sentinel_reclassified_2.tif"))
```

For later use I also need to specify the land use classes in the raster

```{r}
classes <- c("grey", "green" ,"crop", "fallow", "barren", "water")
```

### Import site locations

I will build up the points from scratch, since I had issues while trying to directly import shape files from QGIS. 

All the information I need is within my "site_info.csv" file. All site related information are stored inside (location, ids, name, etc.). I use a classical data.frame and set the site_ids as rownames, this works better with the spatial packages than a tidy workflow.  

```{r results='hide', message=FALSE}
site_info <- read.csv("20190515_site_infos.csv", sep = ";",dec=".")
rownames(site_info) = site_info$site_id
```

Next I create my spatial point data frame. For that I need to merge the point coordinates with some additional point data (like the name, transect, area). Furthermore I have to specify the Coordinate system (WGS84 /UTM zone 43n).

```{r}
CountPts <- SpatialPointsDataFrame(site_info[, 9:10], # the coordinates
                                   site_info[, 1:6], # the data
                                   proj4string = CRS(
                                     "+proj=utm +zone=43 +datum=WGS84 +units=m +no_defs+ellps=WGS84+towgs84=0,0,0")) # coordinate system
                                     
```

### Plot raster and points

Just as a rough overview I want to plot the raster and my site locations. The resulting map is very small - but nevertheless you can clearly see the north and the south transect. Note the two sites in the south which are outside the transects eastern boundary.

```{r}
plot(bgl)
points(CountPts)
```


### Create buffers

I want to analyse the landscape around my sites within circular buffers of different radii. To do that I first have to create the buffer polygons. I will use radii of 100, 500, 1000 and 1500 m around the site points. Since some buffers overlap I have to specfiy `dissolve = FALSE`.

```{r}
# Create buffer polygons
bf100_ply   <- buffer(CountPts, width = 100, dissolve = F)
bf500_ply   <- buffer(CountPts, width = 500, dissolve = F)
bf1000_ply  <- buffer(CountPts, width = 1000, dissolve = F)
bf1500_ply  <- buffer(CountPts, width = 1500, dissolve = F)
```

Next I can use the created polygons to clip the large landscape raster. Basicially, I end up with 36 little landscapes for each of my 4 buffer distances. They are saved in a list object.

```{r warning=FALSE}
# Cut raster by buffers
bf100_ras <- list()
for(i in 1:36) {
  bf100_ras[i] <- crop(bgl, bf100_ply[i,])
  bf100_ras[i] <- mask(bf100_ras[[i]], bf100_ply[i,])
}


bf500_ras <- list()
for(i in 1:36) {
  bf500_ras[i] <- crop(bgl, bf500_ply[i,])
  bf500_ras[i] <- mask(bf500_ras[[i]], bf500_ply[i,])
}

bf1000_ras <- list()
for(i in 1:36) {
  bf1000_ras[i] <- crop(bgl, bf1000_ply[i,])
  bf1000_ras[i] <- mask(bf1000_ras[[i]], bf1000_ply[i,])
}

bf1500_ras <- list()
for(i in 1:36) {
  bf1500_ras[i] <- crop(bgl, bf1500_ply[i,])
  bf1500_ras[i] <- mask(bf1500_ras[[i]], bf1500_ply[i,])
}
```

Finally, I still have to specify the names of my 36 little landscapes - the site-ids, of course.

```{r}
# Set the names of the rasters in the list (my site ids)
names(bf100_ras)  = CountPts@data$site_id
names(bf500_ras)  = CountPts@data$site_id
names(bf1000_ras) = CountPts@data$site_id
names(bf1500_ras) = CountPts@data$site_id
```

Ok, just out of fun (and to get an impression): let us plot the first 4 of my little 36 landscapes for the 500 m radius buffers. If I want I could also change the colors to something more appropriate, but I just go with the default for now.

```{r}
plot(bf500_ras[[1]])
plot(bf500_ras[[2]])
plot(bf500_ras[[3]])
plot(bf500_ras[[4]])
```

### Extract spatial data

Now I have to extract the "data" from my 36 little landscape. Basically I want to know how many pixels (size is 10 x 10 m) belong to each landuse class. With that I can later calculate proportional cover. I can do that with the `extract()` function from the raster package. Note that there is another `extract()` in the tidyr package. 

```{r}
# Extract all pixel values of the raster within buffers around sites/points
bf100_px  <- raster::extract(bgl, CountPts, buffer = 100)
bf500_px  <- raster::extract(bgl, CountPts, buffer = 500)
bf1000_px <- raster::extract(bgl, CountPts, buffer = 1000)
bf1500_px <- raster::extract(bgl, CountPts, buffer = 1500)
```

#### Pixel counts per class

Now I have all the raw pixel values. But I want to count them for each class seperatetly and get a table with the result. For this I set up two functions (following some code I got from Kevin Li). The first function counts the pixel per class and the second takes the results and put them into a table. Both functions need the lapply() command to work, since the previous spatial objects I created are list objects. I will use these functions in the next steps. Note that these functions do not return a tibble but a matrix.

```{r}
# Function: Counts the pixels per class per buffer (needs lapply() to work)
count_class_px <- function(raster_values) {
  levels = 1:length(classes)
  px <- c()
  for(i in 1:length(levels)) {
    px[i] <- sum(raster_values == levels[i], na.rm = T)  
  }
  names(px) = classes
  return(px)
}

# Function: table of class pixel counts
table_class_count <- function(raster_values) {
  px_list <- lapply(raster_values,count_class_px )
  px_mat <- matrix(unlist(px_list), nrow = length(px_list), byrow = T)
  dimnames(px_mat) = list(CountPts@data$site_id, names(px_list[[1]]))
  return(px_mat)
}
```

#### Proportional cover per class

Next to get basic pixel counts I also want to have the proportions of each class. For this I slightly tweak the functions above.

```{r}
# Function: Percentage of each class per buffer (needs lapply() to work)
perc_class_px <- function(raster_values) {
  levels = 1:length(classes)
  px <- c()
  for(i in 1:length(levels)) {
    px[i] <- sum(raster_values == levels[i], na.rm = T)  
  }
  perc_px <- 100 * px / (sum(px))
  names(perc_px) = classes
  return(perc_px)
}

# Function: table of class percentages
table_class_perc <- function(raster_values) {
  perc_list <- lapply(raster_values, perc_class_px)
  perc_mat <- matrix(unlist(perc_list), nrow = length(perc_list), byrow = T)
  dimnames(perc_mat) = list(CountPts@data$site_id, names(perc_list[[1]]))
  return(perc_mat)
}
```

### Landscape composition

Ok great, with the above functions I can now calculate the landscape compositon, i.e. the proportional share of each landuse class per indiviual buffer/radius.

```{r}
composition <- cbind(
  table_class_perc(bf100_px),
  table_class_perc(bf500_px),
  table_class_perc(bf1000_px),
  table_class_perc(bf1500_px)
)

colnames(composition) = c(
  paste(classes, rep("_100",5), sep = ""),
  paste(classes, rep("_500",5), sep = ""),
  paste(classes, rep("_1000",5), sep = ""),
  paste(classes, rep("_1500",5), sep = "")
)

head(composition)
```

### Landscape diversity

To calculate landscape diversity indicies I use the `diversity()` function of the vegan package. This function needs to be called on a community matrix with sites as rows and the counts of the landuse classes in the columns. Now, I can use the pixel count function `table_class_count()` I created above.

```{r}
diversity <- cbind(diversity(table_class_count(bf100_px)),
                   diversity(table_class_count(bf100_px), "simpson"),
                   diversity(table_class_count(bf500_px)),
                   diversity(table_class_count(bf500_px), "simpson"),
                   diversity(table_class_count(bf1000_px)),
                   diversity(table_class_count(bf1000_px), "simpson"),
                   diversity(table_class_count(bf1500_px)),
                   diversity(table_class_count(bf1500_px), "simpson"))
                   
colnames(diversity) = paste(rep(c("shannon", "simpsons"),4), 
                           c("_100","_100", "_500", "_500","_1000","_1000",
                             "_1500", "_1500"), sep ="")

head(diversity)
```

### Landscape configuration

For the landscape configuration metrics I use functions of the landscapemetrics package. These functions return a tibble and the calculated values are found in the last column (...$value). All metrics work on the landscape level, so they return one value for each of my sites/landscape. See the corresponding help pages for the used function for a detailed discription of each of the metrics.

But before I can start I need to omit the two sites for which I still miss the landscape data. The landscapemetric package does not like NAs.

```{r}
bf100_ras[c(9:10)]  <- NULL
bf500_ras[c(9:10)]  <- NULL
bf1000_ras[c(9:10)] <- NULL
bf1500_ras[c(9:10)] <- NULL
```

```{r}
configuration <- tibble(
  mn_p_area_100  = lsm_l_area_mn(bf100_ras)$value, # mean patch area
  mn_p_area_500  = lsm_l_area_mn(bf500_ras)$value,
  mn_p_area_1000 = lsm_l_area_mn(bf1000_ras)$value,
  mn_p_area_1500 = lsm_l_area_mn(bf1500_ras)$value,
  mn_para_100  = lsm_l_para_mn(bf100_ras)$value, # mn paraemter-to-area-ratio
  mn_para_500  = lsm_l_para_mn(bf500_ras)$value,
  mn_para_1000 = lsm_l_para_mn(bf1000_ras)$value,
  mn_para_1500 = lsm_l_para_mn(bf1500_ras)$value,
  mn_shp_100  = lsm_l_shape_mn(bf100_ras)$value, # mn shape index
  mn_shp_500  = lsm_l_shape_mn(bf500_ras)$value,
  mn_shp_1000 = lsm_l_shape_mn(bf1000_ras)$value,
  mn_shp_1500 = lsm_l_shape_mn(bf1500_ras)$value,
  edge_den_100  = lsm_l_ed(bf100_ras)$value, # edge density
  edge_den_500  = lsm_l_ed(bf500_ras)$value,
  edge_den_1000 = lsm_l_ed(bf1000_ras)$value,
  edge_den_1500 = lsm_l_ed(bf1500_ras)$value,
  patch_den_100  = lsm_l_pd(bf100_ras)$value, # patch density
  patch_den_500  = lsm_l_pd(bf500_ras)$value,
  patch_den_1000 = lsm_l_pd(bf1000_ras)$value,
  patch_den_1500 = lsm_l_pd(bf1500_ras)$value,
  no_patches_100 = lsm_l_np(bf100_ras)$value, # number of patches
  no_patches_500 = lsm_l_np(bf500_ras)$value,
  no_patches_1000 = lsm_l_np(bf1000_ras)$value,
  no_patches_1500 = lsm_l_np(bf1500_ras)$value
)

head(configuration)
```

In order to have the same number of rows as the other results (to merge them later) I need to re-add the two missing sites as empty rows.

```{r}
# Insert 2 empty rows at positon 9 and 10; I deleted them earlier
configuration <- insertRows(configuration, c(9,10))
```

### Distance to a central point

Since my sites are along an urbanization gradient I would like to know how far they are away from a central point within the city. I specified this point via Qgis. I choose the "Anil Kumble Circle" on the "MG road" as central point. 

```{r results='hide'}
centralP <- readOGR(dsn=".","central_point")
```

I want to get a distance matrix which list the distance from all sites. Kevin Li quickly came up with the following function.

```{r}
disMat <- function(center,points){
  difx <- points[, 1] - center[, 1] # difference in x
  dify <- points[, 2] - center[, 2] # difference in y
  c2   <- apply(cbind(difx, dify), 1, function(r) sqrt(r[1]^2 + r[2]^2))  # apply Pythagorean theorem
  return(c2)
}
```

Now it is super simple to calculate the distance matrix. I also divide by 1000 to get the distances in km.

```{r}
distance <- disMat(centralP@coords, CountPts@coords) / 1000
distance
```

### Create the final output table

Alright, now I can combine all my results/functions into a final table. This table has 36 rows for each of my sites.

```{r}
output <- cbind(
  site_id = site_info$site_id,
  distance,
  composition, 
  configuration,
  diversity)
```

Lastly, I can export my output table. I use R's own .rds formart, which I find easy to work with. csv files tend to have issues when imported (i.e. numbers to factors)

```{r}
write_rds(output, "birds_landscape.rds")
```

All done! Hurray!
